
\newcommand{\hlbox}[3]{
\begin{mdframed}[backgroundcolor=#2!50]

\rightline{ \footnotesize \emph{#3} }

\centering
#1
\end{mdframed}
}

\newcommand{\tbd}[1]{{\color{violet} #1\footnotesize ?}}
\newcommand{\todo}[1]{{\color{todo} #1\footnotesize !}}

\newcommand{\rquestion}[1]{\hlbox{#1}{yellow}{Research Question}}

\newcommand{\eg}{\textit{e.g.}}
\newcommand{\ie}{\textit{i.e.}}
\newcommand{\etc}{\textit{etc.}}
\newcommand{\perse}{\textit{per se}}

\newcommand{\phd}{Ph.D.}

\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\lang}[1]{\textsc{#1}}
\newcommand{\host}[1]{\textsl{#1}}

\newcommand{\api}{API}

\newcommand{\java}{\lang{Java}}
\newcommand{\cs}{\lang{C\#}}
\newcommand{\cpp}{\lang{C++}}
\newcommand{\smalltalk}{\lang{Smalltalk}}
\newcommand{\javascript}{\lang{JavaScript}}
\newcommand{\haskell}{\lang{Haskell}}
\newcommand{\ql}{\lang{QL}}
\newcommand{\fortran}{\lang{Fortran}}
\newcommand{\cobol}{\lang{Cobol}}
\newcommand{\pascal}{\lang{Pascal}}
\newcommand{\apl}{\lang{APL}}
\newcommand{\basic}{\lang{Basic}}
\newcommand{\swift}{\lang{Swift}}
\newcommand{\php}{\lang{PHP}}

\newcommand{\jvm}{JVM}
\newcommand{\jvmti}{JVMTI}
\newcommand{\jni}{JNI}
\newcommand{\lilith}{Lilith}

\newcommand{\instanceof}{\code{instanceof}}
\newcommand{\throw}{\code{throw}}
\newcommand{\smu}{\code{sun.misc.Unsafe}}
\newcommand{\cce}{\code{ClassCastException}}
\newcommand{\npe}{\code{NullPointerException}}
\newcommand{\setjmp}{\code{setjmp}}
\newcommand{\longjmp}{\code{longjmp}}

\newcommand{\asm}{ASM}
\newcommand{\jnif}{JNIF}

\newcommand{\eval}{\code{eval}}

\newcommand{\github}{\host{GitHub}}
\newcommand{\gitlab}{\host{GitLab}}
\newcommand{\bitbucket}{\host{Bitbucket}}
\newcommand{\sourceforge}{\host{SourceForge}}
\newcommand{\maven}{\host{Maven}}
\newcommand{\mavencentral}{\host{Maven Central}}
\newcommand{\npm}{\host{npm}}
\newcommand{\boa}{\host{Boa}}
\newcommand{\candoia}{\host{Candoia}}
\newcommand{\lgtm}{\host{lgtm}}
\newcommand{\rascal}{\host{Rascal}}

\begin{committee}
  \advisor{Prof.}{Matthias Hauswirth}{\USI, Switzerland}
  \coadvisor{Prof.}{Nathaniel Nystrom}{\USI, Switzerland}
  \internalmember{Prof.}{\tbd{Walter Binder}}
  \internalmember{Prof.}{\tbd{Antonio Carzaniga}}
  \internalmember{Prof.}{\tbd{Gabriele Bavota}}
  \internalmember{Prof.}{\tbd{Patrick Eugster}}
  \externalmember{Prof.}{\tbd{Hridesh Rajan}}{\tbd{Iowa State University, United States}}
  \externalmember{Prof.}{\tbd{Tobias Wrigstad}}{\tbd{Uppsala University, Sweden}}
  \externalmember{Prof.}{\tbd{Stefan Hanenberg}}{\tbd{University Duisburg-Essen, Germany}}
\end{committee}

\phddirector{Prof.}{Walter Binder}
\phddirector{Prof.}{Michael Bronstein}

\abstract{
Programming languages offer a wide range of features that aim to improve programmers productivity.
However, to better drive the future evolution of any programming language, we believe it is paramount to have a thorough understanding of how these features are actually being used in real codebases.

Understanding how developers make use of language features can be helpful to a broad audience besides language designers.
It can aid tool builders to make more realistic assumptions;
researchers to improve the state-of-the-art; and
developers to implement more efficient and effective solutions by providing them best practices.

In this proposal, we target four specific \java{} features, namely, \emph{casting}, \emph{reflection}, \emph{exception handling} and the \emph{unsafe} \api{}.
We plan to divise language and \api{} usage patterns at large-scale to properly assess this broad audience.
We hope that having a better understanding on how these features are used, we can make informed decisions for these driving forces.
}

\usepackage{tikz}
\usepackage{listings}
\usepackage{color}
\usepackage{cleveref}
\usepackage{mdframed}
\crefname{section}{\S}{faafd}
